Handling missing data in smartphone location logs

### 2017-12-15 thesisReportApa.Rmd
forgotten commit

keywords          : "Missing Data, Measurement Bias, GPS, Human Mobility"
wordcount         : "2205"

@@

There is no golden standard for dealing with missing data in GPS logs [@barnett_inferring_2016]. Importantly, spatiotemporal data measurements are often correlated in time and space. This means that common methods, such as mean imputation, are unsuitable.  For example, imagine an individual who splits almost all her time between work and home. Suppose she spends a small amount of time commuting between the two along a circular path. Using mean imputation to estimate her missing coordinates, we impute her to be at the midpoint between home and work.  She has never and will never be there! Worryingly, there is little transparency on how researchers deal with missing data [ @jankowska_framework_2015].

@@

Social scientists are most interested in aggregating spatiotemporal data to more socially relevant information, such as distance travelled. As we discussed earlier, aggregations without data processing can be highly biased. However, as an example we calculate time spent at home of the user for each day of the week in the month of February (Figure \@ref(fig:aggrePlot)). Between individuals, time spent at home has been found to be a reliable predictor of extraversion [@harari_using_2016].

```{r, aggrePlot,echo = FALSE,fig.cap="Proportion of time spent at home in February 2017. We estimate this by calculating the mean lattitude and longitude for every 5 minute time period in the month. The user is coded as at home if within 250 meters of the home coordiantes. Missing values are filled in by the previous observed value. Behavioural trends are evident, the user spends more time at home on weekends than on weekdays. Moreover, the variance is greater on Saturday, due to travel.",out.width = '100%'}
include_graphics("img/timeUse.png")
```

@@

Analyses were performed using R and a multitude of other statistical packages [@base;@ggplot2;@dplyr;@sp1;@sp2].

@@

Three imputation methods were selected in order to cover a wide range of techniques applied in the literature:

>>>>

keywords          : "Missing Data, Measurement Bias, GPS, Human Mobility"
wordcount         : "2297"

@@

There is no golden standard for dealing with missing data in GPS logs [@barnett_inferring_2016]. Importantly, spatiotemporal measurements are often correlated in time and space. This means that common methods, such as mean imputation, are unsuitable.  For example, imagine an individual who splits almost all her time between work and home. Suppose she spends a small amount of time commuting between the two along a circular path. Using mean imputation to estimate her missing coordinates, we impute her to be at the midpoint between home and work.  She has never and will never be there! Worryingly, there is little transparency on how researchers deal with missing data [ @jankowska_framework_2015].

@@

Social scientists are most interested in aggregating spatiotemporal data to more socially relevant information. For instance, @palmius_detecting_2017 calculate several measures to predict bipolar depression, such as the  total distance travelled, the amount of time spent at home, travelling and the regularity of diurnal movement. As an example we calculate two of these measures: time spent at home and distance travelled for the month of February. Figure \@ref(fig:aggrePlot) shows how the aggregate measures are influenced by the noise in the measurements and the missingness. In the case of distance travelled, simply filtering out the most inaccurate measures leads to significantly lower estimates. This phenomenon is mapped out on Figure \@ref(fig:accuracyPlot) . On the other hand, the time spent at home would be greatly influenced at times by the imputation method, as several days with long periods of missing data.

```{r, aggrePlot,echo = FALSE,fig.cap="Proportion of time spent at home  and distance travelled in February 2017. We estimate time spent at home by calculating the mean lattitude and longitude for every 5 minute time period in the month. The user is coded as at home if within 250 meters of the home coordiantes. One can see that several days have long periods with missing data. More accurate estimates can be reached by imputing the missing measurements. As for distance travelled, it is evident that although the behavioural trends remain the same, the absolute estimate of distance travelled varies depending on the filter.",out.width = '100%'}
include_graphics("img/aggPlot.png")
```

@@

Analyses were performed using R and a multitude of other statistical packages [@dplyr;@ggplot2;@base;@ggthemes;@sp1;@sp2;@citr;@tidyr;@leaflet;@tibbletime;@padr;@bibtex;@gridExtra;@scales;@rstudio].

@@

Four imputation methods were selected in order to cover a wide range of techniques applied in the literature:

### 2017-12-15 thesisReportApa2.tex
forgotten commit

    \indent Word count: 2205
  }

@@

Within behavioural sciences, researchers have used GPS data to
investigate wideranging topics. For example, Zenk, Schulz, and
Odoms-Young (2009) investigate the effects of the food environment on
eating patterns. G. M. Harari et al. (2016) look at the movement
correlates of personality, finding that extrovers individuals spend less
time at home than introverts. Wang, Harari, Hao, Zhou, and Campbell
(2015) looks at how academic performance is affected by movement
patterns. Palmius et al. (2017) use mobility patterns to predict bipolar
depression.

@@

Missing data is a pervasive issue as it can arise due to several
reasons. Technical reasons include signal loss, battery failure and
device failure. Behavioural reasons include leaving the phone at home or
switching the device off. As a result, secondary logs often contain wide
temporal gaps with no measurements. For instance, several research
groups studying mental health report missing data rates between 30\% to
50\% (Grünerbl et al., 2015; Palmius et al., 2017; Saeb et al., 2015).
Other researchers report similar trends in different fields (e.g. G. M.
Harari et al., 2016; Jankowska, Schipperijn, \& Kerr, 2015).

There is no golden standard for dealing with missing data in GPS logs
(Barnett \& Onnela, 2016). Importantly, spatiotemporal data measurements
are often correlated in time and space. This means that common methods,
such as mean imputation, are unsuitable. For example, imagine an
individual who splits almost all her time between work and home. Suppose
she spends a small amount of time commuting between the two along a
circular path. Using mean imputation to estimate her missing
coordinates, we impute her to be at the midpoint between home and work.
She has never and will never be there! Worryingly, there is little
transparency on how researchers deal with missing data (Jankowska et
al., 2015).

The accuracy of smartphone location measurements is substantially lower
than that of professional GPS trackers. Android phones collect location
information through a variety of methods. Other than GPS measurements,
Androids use less-accurate heuristics such as WiFi access points and
cellphone triangulation. Different methods are used because of
computational and battery constraints. GPS is the most energy consuming
sensor on most smartphones (M. Y. Chen et al., 2006; LaMarca et al.,
2005). In professional GPS trackers less than 80\% of measurements fall
within 10 meters of the true location. GPS measures are most inaccurate
in dense urban locations and indoors (S. Duncan et al., 2013;
Schipperijn et al., 2014). Unfortunately for researchers, this is where
people in the developed world spend most of their time.

@@

Social scientists are most interested in aggregating spatiotemporal data
to more socially relevant information, such as distance travelled. As we
discussed earlier, aggregations without data processing can be highly
biased. However, as an example we calculate time spent at home of the
user for each day of the week in the month of February (Figure
\ref{fig:aggrePlot}). Between individuals, time spent at home has been
found to be a reliable predictor of extraversion (G. M. Harari et al.,
2016).

\begin{figure}
\includegraphics[width=1\linewidth]{img/timeUse} \caption{Proportion of time spent at home in February 2017. We estimate this by calculating the mean lattitude and longitude for every 5 minute time period in the month. The user is coded as at home if within 250 meters of the home coordiantes. Missing values are filled in by the previous observed value. Behavioural trends are evident, the user spends more time at home on weekends than on weekdays. Moreover, the variance is greater on Saturday, due to travel.}\label{fig:aggrePlot}
\end{figure}

@@

In climate or geological research spatiotemporal imputation methods are
often used. For instance, the CUTOFF method estimates missing values
using the nearest observed neighbours (Feng, Nowak, O'Neill, \& Welsh,
2014 ). The authors illustrate their example using rainfall data from
gauging stations across Australia. Similarly, Z. Zhang et al. (2017) use
a variety of machine learning methods to impute missing values. The
example provided relates to underground water data. Generally these
models assume fixed measurement stations (such as rainfall gauging
stations).

@@

Analyses were performed using R and a multitude of other statistical
packages (Bivand, Pebesma, \& Gomez-Rubio, 2013; E. J. Pebesma \&
Bivand, 2005; R Core Team, 2017; Wickham, 2009; Wickham \& Francois,
2016).

@@

Each imputation method used each of the three filtering methods as an
input. Thus in the end we end up with eight methods to evaluate, three
for each filtering method as well as four for each imputation method.

@@

\hypertarget{refs}{}
\hypertarget{ref-barnett_inferring_2016}{}
Barnett, I., \& Onnela, J.-P. (2016). Inferring Mobility Measures from
GPS Traces with Missing Data. \emph{arXiv:1606.06328 {[}Stat{]}}.
Retrieved from \url{http://arxiv.org/abs/1606.06328}

@@

@@

@@

@@

@@

\hypertarget{ref-dplyr}{}
Wickham, H., \& Francois, R. (2016). \emph{Dplyr: A grammar of data
manipulation}. Retrieved from
\url{https://CRAN.R-project.org/package=dplyr}

>>>>

    \indent Word count: 2297
  }

@@

Within behavioural sciences, researchers have used GPS data to
investigate wideranging topics. For example, Zenk, Schulz, and
Odoms-Young (2009) investigate the effects of the food environment on
eating patterns. Harari et al. (2016) look at the movement
correlates of personality, finding that extroverted individuals spend less
time at home than introverts. Wang, Harari, Hao, Zhou, and Campbell
(2015) looks at how academic performance is affected by movement
patterns. Palmius et al. (2017) use mobility patterns to predict bipolar
depression.

@@

Missing data is a pervasive issue as it can arise due to several
reasons. Technical reasons include signal loss, battery failure and
device failure. Behavioural reasons include leaving the phone at home or
switching the device off. As a result, secondary logs often contain wide
temporal gaps with no measurements. For instance, several research
groups studying mental health report missing data rates between 30\% to
50\% (Grünerbl et al., 2015; Palmius et al., 2017; Saeb et al., 2015).
Other researchers report similar trends in different fields (e.g.
Harari et al., 2016; Jankowska, Schipperijn, \& Kerr, 2015).

There is no golden standard for dealing with missing data in GPS logs
(Barnett \& Onnela, 2016). Importantly, spatiotemporal measurements are
often correlated in time and space. This means that common methods, such
as mean imputation, are unsuitable. For example, imagine an individual
who splits almost all her time between work and home. Suppose she spends
a small amount of time commuting between the two along a circular path.
Using mean imputation to estimate her missing coordinates, we impute her
to be at the midpoint between home and work. She has never and will
never be there! Worryingly, there is little transparency on how
researchers deal with missing data (Jankowska et al., 2015).

The accuracy of smartphone location measurements is substantially lower
than that of professional GPS trackers. Android phones collect location
information through a variety of methods. Other than GPS measurements,
Androids use less-accurate heuristics such as WiFi access points and
cellphone triangulation. Different methods are used because of
computational and battery constraints. GPS is the most energy consuming
sensor on most smartphones (Chen et al., 2006; LaMarca et al.,
2005). In professional GPS trackers less than 80\% of measurements fall
within 10 meters of the true location. GPS measures are most inaccurate
in dense urban locations and indoors (Duncan et al., 2013;
Schipperijn et al., 2014). Unfortunately for researchers, this is where
people in the developed world spend most of their time.

@@

Social scientists are most interested in aggregating spatiotemporal data
to more socially relevant information. For instance, Palmius et al.
(2017) calculate several measures to predict bipolar depression, such as
the total distance travelled, the amount of time spent at home,
travelling and the regularity of diurnal movement. As an example we
calculate two of these measures: time spent at home and distance
travelled for the month of February. Figure \ref{fig:aggrePlot} shows
how the aggregate measures are influenced by the noise in the
measurements and the missingness. In the case of distance travelled,
simply filtering out the most inaccurate measures leads to significantly
lower estimates. This phenomenon is mapped out on Figure
\ref{fig:accuracyPlot} . On the other hand, the estimated time spent at
home on any given day can depend greatly on the imputation method due to
the extent of missing data.

\begin{figure}
\includegraphics[width=1\linewidth]{img/aggPlot} \caption{Proportion of time spent at home  and distance travelled in February 2017. We estimate time spent at home by calculating the mean lattitude and longitude for every 5 minute time period in the month. The user is coded as at home if within 250 meters of the home coordiantes. One can see that several days have long periods with missing data. More accurate estimates can be reached by imputing the missing measurements. As for distance travelled, it is evident that although the behavioural trends remain the same, the absolute estimate of distance travelled varies depending on the filter.}\label{fig:aggrePlot}
\end{figure}

@@

In climate or geological research spatiotemporal imputation methods are
often used. For instance, the CUTOFF method estimates missing values
using the nearest observed neighbours (Feng, Nowak, O'Neill, \& Welsh,
2014 ). The authors illustrate their example using rainfall data from
gauging stations across Australia. Similarly, Zhang et al. (2017) use
a variety of machine learning methods to impute missing values. The
example provided relates to underground water data. Generally these
models assume fixed measurement stations (such as rainfall gauging
stations).

@@

Analyses were performed using R and a multitude of other statistical
packages (Arnold, 2017; Auguie, 2017; Aust, 2016; Bivand, Pebesma, \&
Gomez-Rubio, 2013; Cheng, Karambelkar, \& Xie, 2017; Francois, 2017; E.
J. Pebesma \& Bivand, 2005; R Core Team, 2017; RStudio Team, 2015;
Thoen, 2017; Vaughan \& Dancho, 2017; Wickham, 2009, 2017; Wickham \&
Henry, 2017; Wickham, Francois, Henry, \& Muller, 2017).

@@

Each imputation method used each of the three filtering methods as an
input. Thus in the end we have seven methods to evaluate: three
for each filtering method as well as four for each imputation method.

@@

\hypertarget{refs}{}
\hypertarget{ref-ggthemes}{}
Arnold, J. B. (2017). \emph{Ggthemes: Extra themes, scales and geoms for
'ggplot2'}. Retrieved from
\url{https://CRAN.R-project.org/package=ggthemes}

\hypertarget{ref-gridExtra}{}
Auguie, B. (2017). \emph{GridExtra: Miscellaneous functions for ``grid''
graphics}. Retrieved from
\url{https://CRAN.R-project.org/package=gridExtra}

\hypertarget{ref-citr}{}
Aust, F. (2016). \emph{Citr: 'RStudio' add-in to insert markdown
citations}. Retrieved from \url{https://CRAN.R-project.org/package=citr}

\hypertarget{ref-barnett_inferring_2016}{}
Barnett, I., \& Onnela, J.-P. (2016). Inferring Mobility Measures from
GPS Traces with Missing Data. \emph{arXiv:1606.06328 {[}Stat{]}}.
Retrieved from \url{http://arxiv.org/abs/1606.06328}

@@

\hypertarget{ref-leaflet}{}
Cheng, J., Karambelkar, B., \& Xie, Y. (2017). \emph{Leaflet: Create
interactive web maps with the javascript 'leaflet' library}. Retrieved
from \url{https://CRAN.R-project.org/package=leaflet}

@@

\hypertarget{ref-bibtex}{}
Francois, R. (2017). \emph{Bibtex: Bibtex parser}. Retrieved from
\url{https://CRAN.R-project.org/package=bibtex}

@@

\hypertarget{ref-rstudio}{}
RStudio Team. (2015). \emph{RStudio: Integrated development environment
for r}. Boston, MA: RStudio, Inc. Retrieved from
\url{http://www.rstudio.com/}

@@

\hypertarget{ref-padr}{}
Thoen, E. (2017). \emph{Padr: Quickly get datetime data ready for
analysis}. Retrieved from \url{https://CRAN.R-project.org/package=padr}

\hypertarget{ref-tibbletime}{}
Vaughan, D., \& Dancho, M. (2017). \emph{Tibbletime: Time aware
tibbles}. Retrieved from
\url{https://CRAN.R-project.org/package=tibbletime}

@@

\hypertarget{ref-scales}{}
Wickham, H. (2017). \emph{Scales: Scale functions for visualization}.
Retrieved from \url{https://CRAN.R-project.org/package=scales}

\hypertarget{ref-tidyr}{}
Wickham, H., \& Henry, L. (2017). \emph{Tidyr: Easily tidy data with
'spread()' and 'gather()' functions}. Retrieved from
\url{https://CRAN.R-project.org/package=tidyr}

\hypertarget{ref-dplyr}{}
Wickham, H., Francois, R., Henry, L., \& Muller, K. (2017). \emph{Dplyr:
A grammar of data manipulation}. Retrieved from
\url{https://CRAN.R-project.org/package=dplyr}
